---
title: "HW #1 - Portfolio Construction"
author: "Ivan Rodriguez"
output:
  pdf_document:
    toc: true
    highlight: default
---


---

#Data Notes

The data used in the analysis is the Fama French portfolio and factors database from WRDS. The data are monthly value weighted return observations on the 6 portfolio constructed by Fama and French formed on size and book to market. The time period of the data is from the end of March 2000 to the end of July 2014. The portfolio variable names are as follows:

- SMLO (small size and low book-to-market ratio equities),
- SMME (small size and medium book-to-market ratio equities),
- SMHI (small size and high book-to-market ratio equities),
- BILO (big size and low book-to-market ratio equities),
- BIME (big size and medium book-to-market ratio equities),
- BIHI (big size and high book-to-market ratio equities)

Here is a small sample of the data that was used:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
options(digits=5, width = 90)
#Set working directory
setwd("~/Documents/projects/class/FIN7809_theoryII/HW1")

#Loading in data file
data.FF1 <- read.csv(file = "FF1.csv", head = TRUE, sep = ",")
data.FF2 <- read.csv(file = "FF2.csv", head = TRUE, sep = ",")

#Loading dplyr package to transform raw data
library("dplyr")
library("ggplot2")
library("GGally")
library("reshape2")

### DATA CLEANING ###
data.1 <- tbl_df(data.FF1) #Simplify data frame using tbl_df
data.2 <- tbl_df(data.FF2) #Simplify data frame using tbl_df

#Select relevant variables
data1 <- select(data.1, date,smlo_vwret,smme_vwret,smhi_vwret,bilo_vwret,bime_vwret,bihi_vwret)
data2 <- select(data.2, date = dateff, rf)

#Create final data set of interest
data <- inner_join(data1, data2)
data.portfolio <- select(data, -date, -rf)

n <- ncol(data.portfolio) #number of assets

head(data, n=4)
```

#Questions
Before proceeding with answering the homework questions, we need to initialize a few things first, namely the unity vector. The unity vector $\boldsymbol{\iota}$ is the `r n` x 1 vector of ones.

```{r}
i <- 1 + numeric(n)
```

##1a. The 6x1 vector of mean returns, $\boldsymbol{\mu}$.
The vector $\boldsymbol{\mu}$ is the `r n` x 1 vector of the mean returns.
```{r}
mu <- sapply(data.portfolio, mean)
mu
```

##1b. The 6x6 covariance matrix, $\boldsymbol{\Omega}$.
The matrix $\boldsymbol{\Omega}$ is the `r n` x `r n` matrix whose elements are the covariances between the portfolios
```{r}
V <- cov(data.portfolio)
V
```

##1c. Find the intermediate values of A, B, C, and D.
The intermediate values, which will make the later computation easier, are found through solving the original optimization problem and some matrix algebraic manipulation. 

###Intermediate Value A
The value of A is found to be $A = \boldsymbol{\iota}' \boldsymbol{\Omega}^{-1} \boldsymbol{\mu}$.

```{r}
A <- t(i) %*% solve(V) %*% mu
A
```

###Intermediate Value B
The value of B is found to be $B = \boldsymbol{\mu}' \boldsymbol{\Omega}^{-1} \boldsymbol{\mu}$.

```{r}
B <- t(mu) %*% solve(V) %*% mu
B
```

###Intermediate Value C
The value of C is found to be $C = \boldsymbol{\iota}' \boldsymbol{\Omega}^{-1} \boldsymbol{\iota}$.

```{r}
C <- t(i) %*% solve(V) %*% i
C
```

###Intermediate Value D
The value of D is found to be $D = BC - A^2$.

```{r}
D <- B * C - A^2
D
```

##1d. Find the 6x1 vector $\mathbf{g}$
The vector $\mathbf{g}$ is the `r n` x 1 vector which is equal to $\frac{1}{D} \left[ B(\boldsymbol{\Omega}^{-1}\boldsymbol{\iota}) - A(\boldsymbol{\Omega}^{-1}\boldsymbol{\mu}) \right]$.

```{r}
g <- 1/drop(D) * (drop(B) * (solve(V) %*% i) - drop(A) * (solve(V) %*% mu))
g <- g[,1]
g
```

##1e. Find the 6x1 vector $\mathbf{h}$
The vector $\mathbf{h}$ is the `r n` x 1 vector which is equal to $\frac{1}{D} \left[ C(\boldsymbol{\Omega}^{-1}\boldsymbol{\mu}) - A(\boldsymbol{\Omega}^{-1}\boldsymbol{\iota}) \right]$.

```{r}
h <- 1/drop(D) * (drop(C) * (solve(V) %*% mu) - drop(A) * (solve(V) %*% i))
h <- h[,1]
h

```


##1f. Find the globabl minimum variance portfolio, $w_g$
The formula for the weights of the global minimum portfolio is $w_g = \frac{1}{C} \boldsymbol{\Omega}^{-1}\boldsymbol{\iota}$.

```{r}
w.gl <- 1/drop(C) * solve(V) %*% i
w.gl <- w.gl[,1]
w.gl
```

##1g. Find the globabl minimum variance portfolio, $\mu_g$
The formula for the return of the global minimum portfolio is $\mu_g = \frac{A}{C}$.
```{r}
mu.gl <- drop(A) / drop(C)
mu.gl
```

##1h. Find the globabl minimum variance portfolio, $\sigma_g$
The formula for the standard deviationof the global minimum portfolio is $\sigma_g = \frac{1}{C}$.
```{r}
sigma.gl <- 1 / drop(C)
sigma.gl
```

##1i. Find weight for an efficient portfolio with a mean equal to 3.5%, and call this portfolio $p$
The formula for the optimal weight for a portfolio $p$ is $w_p = \mathbf{g} + \mathbf{h}\mu_p$.
```{r}
mu.p <- .035
w.p <- g + h * mu.p
w.p
```

##1j. Find the weights, $\mathbf{w}_{op}$, and the mean , $\mu_{op}$, for portfolio $p$'s zero beta portfolio
The return of the portfolio which is uncorrelated with portfolio $p$ is $\mu_{op} = \frac{D/C^2}{\mu_p - D/C} + \frac{A}{C}$.

```{r}
mu.0p <- -(D/C^2)/(mu.p - A/C) + A/C
mu.0p
```

The weight vector of the portfolio is given by $\mathbf{w}_{op} = \mathbf{g} + \mathbf{h}\mu_{op}$.

```{r}
w.0p <- g + h * mu.0p
w.0p
```

##1k. Find the regression beta of the first portfolio's return with respect to portfolio $p$.
Now, we will SMLO's return's on the portfolio $p$'s return. In order to run the regression, first we have to find the return of our portfolio $p$ over the same time period as SMLO, denoted as $\boldsymbol{\mu}_P$. We simply use the following formula, $\boldsymbol{\mu}_P = \mathbf{X} \mathbf{w}_p$ where $\mathbf{X}$ is our data matrix. 

```{r}
port_vwret <- as.matrix(data.portfolio) %*% w.p
port_vwret <- data.frame(port_vwret)
data <- tbl_df(cbind(data, port_vwret))

#Creating separate regression data table in percent instead of decimals 
data.regression <- select(data, smlo_vwret, port_vwret) * 100
head(data.regression, n=4)
```

Now that we have calculated the necessary returns, we can run our regression, $\boldsymbol{\mu}_{SMLO} = \alpha + \beta\boldsymbol{\mu}_p$.

```{r}
reg <- lm(smlo_vwret ~ port_vwret, data=data.regression)
summary(reg)
```

As we can see, our regression beta is $\beta =$ `r summary(reg)$coef[2,1]`.

##1l. Verify that the average return of portfolio 1 can be expressed exactly as $\mu_{SMLO} = \mu_{op} + \beta (\mu_p - \mu_{op})$
First, let's find the mean of SMLO from our data

```{r}
mu[1]
```

Let's now find the mean using our expression

```{r}
mu.expression <- mu.0p + summary(reg)$coef[2,1] * (mu.p - mu.0p)
mu.expression
```
Does the average return of SMLO equal the expression? `r round(mu.expression, 5) == round(mu[1],5)`!

##1m. Use all 6 portfolios and plot them in mean-standard deviation space. Which portfolio shows the best return per unit of risk?

To plot our points and find the portfolio with the highest return per unit of risk, we first have to find their standard deviations, which we denote as the vector $\boldsymbol{\sigma}$.

```{r}
v <- diag(V)
s <- sqrt(v)
s
```

Now, we can find the portfolio with the largest return per unit of risk

```{r}
retrisk <- mu/s
retrisk
max(retrisk)
```

and plot them in mean-standard deviation space

```{r}
qplot(s, mu) + theme_bw() + xlab("Standard Deviation") +
  ylab("Return") +
  ggtitle("Efficient Frontier and Optimal Portfolio Plot")
```

##1n. Use the matrix based techniques learned in class to derive the mean-variance frontier and plot the 6 portfolios

To plot the mean variance frontier, we first have to find the weights of the optimal weights of any efficient portfolio. We can plot each efficient frontier if we had their return and variance. To find the return, we can use this formula: $\mu_e = \alpha \mu_g + (1-\alpha)\mu_p$. To find the variance, we first have to find the weight of the efficient frontier this using this formula: $w_e = \alpha (\mathbf{g} + \mathbf{h}\mu_g) + \alpha (\mathbf{g} + \mathbf{h}\mu_p)$.

```{r}
#Initialize vectors that are going to be used
alpha <- seq(0,2,.01)
return.eff = NULL
std.eff = NULL

#loop for every alpha and write to our intialized vectors
for (j in 1:length(alpha)) {
  return.eff[j] = alpha[j] * mu.gl + (1 - alpha[j]) * mu.p
  weight.eff = alpha[j] * (g + h*mu.gl) + (1 - alpha[j]) * (g + h*mu.p)
  std.eff[j] = sqrt(diag(t(weight.eff) %*% V %*% weight.eff))
}

#Create data table for graph
frontier.data <- data.frame(cbind(return.eff, std.eff))

#Plot frontier and portfolios
ggplot() + 
  geom_path(data=frontier.data, aes(x=std.eff, y=return.eff)) + 
  geom_point(aes(x=s, y=mu)) + 
  xlab("Standard Deviation") +
  ylab("Return") +
  ggtitle("Efficient Frontier and Portfolio Plots") +
  theme_bw()
```

##1o. Using the average T-Bill rate from the Fama French data, calculate the expected return and standard deviation of the optimal portfolio on the efficient frontier and graph
The average one month T-Bill rate over our time period of interest can be used to find the optimal portfolio.

```{r}
rf <- mean(data$rf)
rf
```
We use this avearge as a proxy for the risk free rate, $r_f =$ `r rf`, we can find the weight of this optimal portfolio $w_t = \frac{1}{\boldsymbol{\iota}'\mathbf{z}}$ where $\mathbf{z} = \boldsymbol{\Omega}^{-1}(\boldsymbol{\mu}-r_f)$.

```{r}
rf <- mean(data$rf)
z <- solve(V) %*% (mu - rf)
z.sum <- t(i) %*% z
w_t <- 1/drop(z.sum) * z
w_t <- w_t[,1]
w_t
```

Using these weights, we can find the expected return and standard deviation of the optimal portfolio.

```{r}
mu.t <- mean(as.matrix(data.portfolio) %*% w_t)
mu.t
std.t <- sqrt(diag(t(w_t) %*% V %*% w_t))
std.t
```

We can plot both of these points and see where this lie on the efficient frontier.

```{r}
#Plot frontier and portfolios
ggplot() + 
  geom_path(data=frontier.data, aes(x=std.eff, y=return.eff)) + 
  geom_point(aes(x=s, y=mu)) + 
  geom_point(aes(x=std.t, y=mu.t), colour="#CC0000") +
  xlab("Standard Deviation") +
  ylab("Return") +
  ggtitle("Efficient Frontier and Optimal Portfolio Plot") +
  theme_bw()
```

##1p. Assume a risk aversion coefficient of $A = 3$ and calculate the proportion of investment in the risky optimal portfolio and the risk free asset.

The proportion invested into the complete portfolio, denoted $y$ is given by the equation $y = \frac{E(r_t) - r_f}{0.5A\sigma_t^2}$.

```{r}
A = 3
y <- (mu.t - rf)/(0.5 * A * std.t^2)
y
```
This means that the amount invested in the optimal portfolio is $y=$ `r y` and the amount invested in the risk free is $1-y =$ `r 1 - y`.
